\documentclass[runningheads,a4paper]{llncs}

\usepackage{amssymb}
\usepackage{amsmath}

\usepackage{mathrsfs}
\usepackage{stmaryrd}

\usepackage{enumitem}
%\usepackage{enumerate}

\usepackage{color}
\usepackage{graphicx}
\usepackage{rotating}
%\usepackage{xparse}
%\usepackage{latex8}
\usepackage{upgreek} 
\usepackage{cmll}
\usepackage{url}
\usepackage{xifthen}% provides \isempty test
\usepackage{multirow}

\setcounter{tocdepth}{3}

\urldef{\mailsa}\path|{acarraro}@dsi.unive.it|
%\urldef{\mailsb}\path||
%\urldef{\mailsc}\path|
\newcommand{\keywords}[1]{\par\addvspace\baselineskip
\noindent\keywordname\enspace\ignorespaces#1}

\input prooftree.sty
\input xy
\xyoption{all}

\makeindex

\begin{document}

\mainmatter  % start of an individual contribution

% first the title is needed
\title{Note del corso di Calcolabilit\`{a} e Linguaggi Formali - Lezione 12}

% a short form should be given in case it is too long for the running head
\titlerunning{Note del corso di Calcolabilit\`{a} e Linguaggi Formali - Lezione 12}

% the name(s) of the author(s) follow(s) next
%
% NB: Chinese authors should write their first names(s) in front of
% their surnames. This ensures that the names appear correctly in
% the running heads and the author index.
%
\author{Alberto Carraro \\ 14 dicembre 2016}
%
\authorrunning{A. Carraro}
% (feature abused for this document to repeat the title also on left hand pages)

% the affiliations are given next; don't give your e-mail address
% unless you accept that it will be published
\institute{IIS Bruno-Franchetti, Mestre (VE)
%\mailsa\\
%\mailsb\\
%\mailsc\\
\url{http://www.dsi.unive.it/~acarraro}
}

%
% NB: a more complex sample for affiliations and the mapping to the
% corresponding authors can be found in the file "llncs.dem"
% (search for the string "\mainmatter" where a contribution starts).
% "llncs.dem" accompanies the document class "llncs.cls".
%

\toctitle{Note del corso di Calcolabilit\`{a} e Linguaggi Formali - Lezione 12}
\tocauthor{A. Carraro}

\input{macro}

\maketitle
%\tableofcontents


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Parity games}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{definition}\label{def:parity-game}
A \emph{parity game} is a tuple $\cG = (V, R, \lambda, C, \Omega)$
\begin{itemize}
\item $V$ is a finite set of states vertices partitioned into two sets $V_V$ (vertices owned by "Verifier") and $V_R$ (vertices owned by "Refuter"),
\item $R \subseteq V \times V$ is the accessibility relation,
\item $C$ is a finite set of labels and $\lambda: V \times C$ is the labelling function,
\item $\Omega: V \to \{1,\ldots, k\}$ is the priority function.
\end{itemize}
\end{definition}

A \emph{play} is a finite or infinite sequence of vertices. A finite play $\pi$ is winning for Verifier if its last vertex  belongs to $V_V$. An infinite play $\pi = v_0, v1, \ldots$ is winning for Verifier if in the sequence $\Omega(\pi) = \Omega(v_0), \Omega(v1), \ldots$ the least priority that occurs infinitely often is even. We let 
\begin{itemize}
\item $Win = \{\pi : \pi \text{ is a winning play for Verifier} \}$, so that $Win \subseteq V^* \cup V^\omega$.
\item $Acc = \{\lambda(\pi) : \pi \in Win \}$, so that $Acc \subseteq C^* \cup C^\omega$.
\end{itemize}

\begin{example}
For a graph $(V, R)$ and a finite set $F \subseteq C$. We can build a game such that  $Acc \cap C^\omega = F^\omega$ by setting:
$\Omega(v) = 0$ for every $v \in F$.
\end{example}

\begin{example}
For a graph $(V, R)$ and a finite set $F \subseteq C$ we let $Reach_F$ be the set of plays $\pi$ with $\pi = v_0,v_1,\ldots \in V^\omega$ and such that there exists $i$ with $\lambda(v_i) \in F$. We can build a game such that  $Acc \cap C^\omega = \lambda(Reach_F)$ by setting:

\end{example}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Nondeterministic tree automata}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{definition}\label{def:NDPTA}
A \emph{nondeterministic parity tree automaton} (NPTA) for $\Sigma$-labelled full binary trees is a tuple $A = (Q, \Sigma, q_I, \delta, \Omega)$ where 
\begin{itemize}
\item $Q$ is a finite set of states,
\item $q_I \in Q$ is the initial state, 
\item $\delta: Q \times \Sigma \to \cP(Q^2)$ is the transition function and 
\item $\Omega: Q \to \{1,\ldots, k\}$ is the priority function.
\end{itemize}
\end{definition}

\begin{remark}
Note that the transition function $\delta: Q \times \Sigma \to \cP(Q^2)$ is by some authors seen as a function
 $\delta: Q \times \Sigma \times \{0,1\} \to \cP(Q)$.
\end{remark}

A \emph{run-tree} of an automaton $A$ over a tree $t$ is a map $\rho: \{0,1\}^* \to Q$ such that 
\begin{enumerate}
\item $\rho(\varepsilon) = q_I$ and
\item for all $u \in \{0,1\}^*$ we have $(\rho(u0), \rho(u1)) \in \delta(\rho(u), t(u))$.
\end{enumerate}

A run-tree $\rho$ of $A$ over $t$ is \emph{accepting} iff in every path $\pi$ of $\rho$ the least priority that occurs infinitely often is even. An automaton $A$ \emph{accepts} a tree $t$ just if there exists an accepting run-tree of $A$ over $t$.

\begin{definition}\label{def:acceptance-parity-game}
Given an NPTA $A = (Q, \Sigma, q_I, \delta, \Omega)$ and a tree $t$ the \emph{acceptance parity game} 
 $\cG_{A,t} = (V_V, V_R, E, v_I, \Omega')$ where 
\begin{itemize}
\item $V_V = \{0,1\}^* \times Q$
\item $V_R = \{0,1\}^* \times Q^2$
\item $v_I = (\varepsilon, q_I)$
\item $E \subseteq (V_V \times V_R) \cup (V_R \times V_V)$ is the smallest relation such that
\begin{itemize}
\item if $(v,q) \in V_V$ and $(q_0, q_1) \in \delta(q, a)$ and $t(v) = a$, then \mbox{$((v,q), (v,(q_0,q_1))) \in E$}
\item if $(v,(q_0,q_1)) \in V_R$, then $((v,(q_0,q_1)), (v0, q_0) ) \in E$ and \mbox{$((v,(q_0,q_1)), (v1, q_1) ) \in E$}
\end{itemize}
\item $\Omega'(v,q) = \Omega(q)$ and $\Omega'(v,(q_0,q_1)) = \mathrm{max}\{\Omega(q_0),\Omega(q_1)\}$
\end{itemize}
\end{definition}

The acceptance parity game of Definition~\ref{def:acceptance-parity-game} can be read as follows:
\begin{itemize}
\item Verifier can exploit the nondeterminism of $A$ by choosing the run-tree
\item Refuter chooses the path
\end{itemize}
Their combined action is headed towards checking that there exists ($\exists$) a run-tree all ($\forall$) of whose paths are accepting. Again Verifier represents ($\exists$) and Refuter represents ($\forall$).

\begin{lemma}\label{lem:acceptance-game}
Verifier has a winning strategy in $\cG_{A,t}$ from vertex $v_I$ iff $t \in L(A)$.
\end{lemma}

\begin{proof}
\noindent ($\Rightarrow$) Suppose Verifier has a winning strategy $\sigma$ in $\cG_{A,t}$ from $v_I$. Define the map $\rho: \{0,1\}^* \to Q$ inductively as 
\begin{itemize}
\item $\rho(\varepsilon) = q_I$\ ;
\item $\rho(v0) = q_0$ if $\rho(v) = q$ and $\sigma(v,q) = (v,(q_0,q_1))$\ ;
\item $\rho(v1) = q_1$ if $\rho(v) = q$ and $\sigma(v,q) = (v,(q_0,q_1))$\ .
\end{itemize}
Clearly $\rho$ is a run-tree of $A$ over $t$. A path $\pi$ of $\rho$ has vertices \\
 $\ldots,(v,q), (v,(q_0,q_1)), (v0, q_0), (v0,(q_0',q_1')), (v01, q_1'),\ldots$ \\ and therefore has priorities\\
 $\ldots,\Omega(q), \mathrm{max}\{\Omega(q_0), \Omega(q_1)\}, \Omega(q_0), \mathrm{max}\{\Omega(q_0'), \Omega(q_1')\}, \Omega(q_1'),\ldots$.\\
The least priority occurring infinitely often is even but since each priority appearing at an even position is $\geq$ than its immediate successor, the subsequence obtained by removing all priorities at even positions is such that the least priority occurring infinitely often is still even. Such a subsequence witnesses an accepting path of $\rho$ and all paths of $\rho$ are obtained this way. Hence $\rho$ is accepting.

\noindent ($\Leftarrow$) Suppose $\rho$ is an accepting run-tree fo $A$ over $t$. Then define the map 
 $\sigma: (V_V \times V_R^*)V_V \to V_R$ as 
\begin{itemize}
\item $\sigma(\alpha (v,q)) = (v,(q_0,q_1))$ iff $\rho(v0) = q_0$ and $\rho(v1) = q_1$.
\end{itemize}
Clearly $\sigma$ is a strategy for Verifier in $\cG_{A,t}$ from $v_0$. A play $\pi$ respecting $\sigma$ has vertices \\
 $\ldots,(v,q), (v,(q_0,q_1)), (v0, q_0), (v0,(q_0',q_1')), (v01, q_1'),\ldots$ \\ and therefore has priorities\\
 $\ldots,\Omega(q), \mathrm{max}\{\Omega(q_0), \Omega(q_1)\}, \Omega(q_0), \mathrm{max}\{\Omega(q_0'), \Omega(q_1')\}, \Omega(q_1'),\ldots$.\\
The subsequence obtained by removing all priorities at even positions is the run-tree $\rho$ and the least priority occurring infinitely often in it is even. Since each priority appearing at an even position is $\geq$ than its immediate successor, the least priority occurring infinitely often in $\pi$ is still even. Such a play is winning for Verifier.
\end{proof}

\begin{definition}\label{def:non-emptiness-parity-game}
Given an NPTA $A = (Q, \Sigma, q_I, \delta, \Omega)$ the \emph{non-emptiness parity game} 
 $\cG_{A} = (V_V, V_R, E, v_I, \Omega')$ where 
\begin{itemize}
\item $V_V = Q$
\item $V_R = \{((q,a), (q_0,q_1)) \st (q_0,q_1) \in \delta(q,a)\}$.
\item $v_I = q_I$
\item $E \subseteq (V_V \times V_R) \cup (V_R \times V_V)$ is the smallest relation such that
\begin{itemize}
\item if $q \in V_V$ and $(q_0, q_1) \in \delta(q, a)$, then $(q, ((q,a),(q_0, q_1))) \in E$
\item if $y = ((q,a),(q_0,q_1)) \in V_R$, then $(y, q_0)\in E$ and $(y,q_1) \in E$
\end{itemize}
\item $\Omega'(q) = \Omega(q)$ and $\Omega'((q,a),(q_0,q_1)) = \Omega(q)$
\end{itemize}
\end{definition}

\begin{lemma}\label{lem:non-emptiness-game}
Verifier has a winning strategy in $\cG_{A}$ from vertex $v_I$ iff $L(A) \neq \emptyset$.
\end{lemma}

\begin{proof}
\noindent ($\Rightarrow$) Suppose Verifier has a winning strategy $\sigma$ in $\cG_{A}$ from $v_I$. Then $\delta(q,a) \neq \emptyset$ for every pair $(q,a) \in Q \times \Sigma$. The solution to the fixpoint equation  
 $\tau(q) = (a,\tau(q_0),\tau(q_1))$ with constraints $\sigma(q) = ((q,a),(q_0, q_1))$ is a tree whose first coordinates
  define a tree $t \in L(A)$.\\

\noindent ($\Leftarrow$) Let $\alpha \in (V_VV_R)^*V_V$ be an initial segment of a play in $\cG_{A}$ from $v_I$. 
 The run-trees $\rho$ of the trees $t \in L(A)$.
Suppose  is an input tree and let $\rho$ be an associated run-tree. 

Let $\bar{t}(\varepsilon) = t(\varepsilon)$ and $\bar{t}(uj) = \bar{t}(u)t(uj)$ for $u \in \{0,1\}^*$ and $j \in \{0,1\}$. 
 Let also $\bar{\rho}(\varepsilon) = \rho(\varepsilon)$ and $\bar{\rho}(uj) = \bar{\rho}(u)\rho(uj)$ for $u \in \{0,1\}^*$ and $j \in \{0,1\}$.

Come si fa a definire la strategia vincente? La run-tree su $t$ potrebbe non visitare tutti gli stati e quindi per gli stati ne non compaiono come faccio ad avere la scelta giusta del successore per la strategia? 

 is a sequence $x_1(y_1,z_1), \ldots,x_n(y_n,z_n)$ with $x_i \in Q$, $y_i \in Q \times \Sigma$ and $z_i \in Q^2$. The trace of $\alpha$ is then $y_1,\ldots, y_n$. Then the map $\sigma: (V_VV_R)^*V_V \to V_R$ given by $\sigma(\alpha q) = $ component-wise first  if $\alpha$ "contains" the string $u = i_1\ldots i_n$ and $t(i_1)\ldots t(i_1\ldots i_n) = a$ and $\rho(i_1)\ldots \rho(i_1\ldots i_n) = q$. 

\end{proof}

\section{Finite automata on infinite words}

A nondeterminitstic \emph{finite automaton} on $\Sigma$-labelled infinite words is a tuple $A = (Q, \Sigma, q_I, \delta)$ where 
\begin{itemize}
\item $Q$ is a finite set of states;
\item $q_I \in Q$ is the initial state; 
\item $\delta: Q \times \Sigma \to \cP(Q)$ is the transition function.
\end{itemize}

A run of a word $\alpha \in \Sigma^\omega$ is a sequence $\varrho \in Q^\Omega$ such that $\varrho(0) = q_I$ and $\varrho(i+1) \in \delta(\varrho(i), \alpha(i))$ for all $i \geq 0$. 

We let $\text{Inf}(\varrho) = \{ q \in Q \st \varrho^{-1}(q) \text{ is infinite} \}$.

\begin{definition}\label{def:NBW}
A \emph{nondeterministic B\"{u}chi automaton} on $\Sigma$-labelled infinite words is a finite automaton $A = (Q, \Sigma, q_I, \delta, F)$ where $F \subseteq Q$ is the set of final states.
 The language of $A$, $L(A)$, is the set of words for which there exists a run $\varrho$ with $\text{Inf}(\varrho) \cap F \neq \emptyset$.
\end{definition}

\begin{definition}\label{def:NMW}
A \emph{nondeterministic M\"{u}ller automaton} on $\Sigma$-labelled infinite words is a finite automaton $A = (Q, \Sigma, q_I, \delta, \Omega)$ where $\Omega = \{F_1,\ldots,F_k\}$ is a finite set with $F_i \subseteq Q$.
 The language of $A$, $L(A)$, is the set of words for which there exists a run $\varrho$ such that $\text{Inf}(\varrho) \in \Omega$.
\end{definition}

\begin{definition}\label{def:NRW}
A \emph{nondeterministic Rabin automaton} on $\Sigma$-labelled infinite words is a finite automaton $A = (Q, \Sigma, q_I, \delta, \Omega)$ where $\Omega = \{(E_1,F_1),\ldots,(E_k,F_k)\}$ is a finite set with $E_i,F_i \subseteq Q$. The language of $A$, $L(A)$, is the set of words for which there exists a run $\varrho$ such that there exists a pair $(E,F) \in \Omega$ such that $\text{Inf}(\varrho) \cap E = \emptyset$ and $\text{Inf}(\varrho) \cap F \neq \emptyset$.
\end{definition}

\begin{definition}\label{def:NPW}
A \emph{nondeterministic Parity automaton} on $\Sigma$-labelled infinite words is a finite automaton $A = (Q, \Sigma, q_I, \delta, \Omega)$ where $\Omega: Q \to \{1,\ldots,k\}$ is the set of priorities (or colors).
 The language of $A$, $L(A)$, is the set of words for which there exists a run $\varrho$ such that the minimum element of $\text{Inf}(\Omega\circ \varrho)$ is even.
\end{definition}

\begin{theorem}\label{thm:Buchi>Muller}
Let $A = (Q, \Sigma, q_I, \delta, F)$ be a nondeterministic B\"{u}chi automaton. Then there exists a nondeterministic M\"{u}ller automaton $A'$ such that $L(A') = L(A)$.
\end{theorem}

\begin{proof}
Define $\Omega = \{G \in \cP(Q) \st G \cap F \neq \emptyset\}$. Then the nondeterministic M\"{u}ller automaton $A'= (Q, \Sigma, q_I, \delta, \Omega)$ has the desired property.
\end{proof}

\begin{theorem}\label{thm:Muller>Buchi}
Let $A = (Q, \Sigma, q_I, \delta, \Omega)$ be a nondeterministic M\"{u}ller automaton. Then there exists a nondeterministic B\"{u}chi automaton $A'$ such that $L(A') = L(A)$.
\end{theorem}

\begin{proof}
Define $Q' = Q \cup \{(G, q, X) \st G \in \Omega, q \in G, X \subseteq G\}$ and $F = \{(G, q, \emptyset) \st G \in \Omega, q \in G\}$. 

Let $q \in Q$ and suppose $p \in \delta(q,a)$. Then also $p \in \delta'(q,a)$ and for each $G \in \Omega$ we have  
$$
\begin{cases}
(G, p, \{q\}) \in \delta'(q, a) & \text{if } \{q\} \subset G \text{ and } p \in G \\
(G, p, \emptyset) \in \delta'(q, a) & \text{if } \{q\} = G \text{ and } p \in G 
\end{cases}
$$
and
$$
\begin{cases}
(G, p, X \cup \{q\}) \in \delta'((G, q, X), a) & \text{if } X \cup \{q\} \subset G \text{ and } p \in G \\
(G, p, \emptyset) \in \delta'((G, q, X), a) & \text{if } X \cup \{q\} = G \text{ and } p \in G 
\end{cases}
$$

The nondeterministic B\"{u}chi automaton $A' = (Q', \Sigma, q_I, \delta', F)$ has the desired property.
\end{proof}

\begin{theorem}\label{thm:Rabin>Muller}
Let $A = (Q, \Sigma, q_I, \delta, \Omega)$ be a nondeterministic Rabin automaton. Then there exists a nondeterministic M\"{u}ller automaton $A'$ such that $L(A') = L(A)$.
\end{theorem}

\begin{proof}
Define $\Omega' = \{G \in \cP(Q) \st \exists (E,F) \in \Omega.\ (G \cap E = \emptyset \wedge G \cap F \neq \emptyset)\}$. Then the nondeterministic M\"{u}ller automaton $A'= (Q, \Sigma, q_I, \delta, \Omega')$ has the desired property.
\end{proof}

\begin{theorem}\label{thm:Parity>Rabin}
Let $A = (Q, \Sigma, q_I, \delta, \Omega)$ be a nondeterministic Parity automaton. Then there exists a nondeterministic Rabin automaton $A'$ such that $L(A') = L(A)$.
\end{theorem}

\begin{proof}
Suppose that $\{0,\ldots,k\}$ is the set of colors for $A$. Let $r = \lfloor \frac{k}{2} \rfloor$ and $\Omega' = \{(E_0,F_0),\ldots,(E_r,F_r)\}$ where
\begin{itemize}
\item $E_i = \{q \in Q \st \Omega(q) < 2i\}$
\item $F_i = \{q \in Q \st \Omega(q) \leq 2i\}$
\end{itemize}
Then the nondeterministic Rabin automaton $A'= (Q, \Sigma, q_I, \delta, \Omega')$ has the desired property.
\end{proof}

\begin{theorem}\label{thm:Buchi>Parity}
Let $A = (Q, \Sigma, q_I, \delta, F)$ be a nondeterministic B\"{u}chi automaton. Then there exists a nondeterministic Parity automaton $A'$ such that $L(A') = L(A)$.
\end{theorem}

\begin{proof}
Let $\{0,1\}$ be the set of colors and define 
$$
\Omega(q) = 
\begin{cases}
0 & \text{if } q \in F \\
1 & \text{if } q \not\in F
\end{cases}
$$
Then the nondeterministic Parity automaton $A'= (Q, \Sigma, q_I, \delta, \Omega)$ has the desired property.
\end{proof}

\begin{theorem}\label{thm:Buchi>Rabin}
Let $A = (Q, \Sigma, q_I, \delta, F)$ be a nondeterministic B\"{u}chi automaton. Then there exists a nondeterministic Rabin automaton $A'$ such that $L(A') = L(A)$.
\end{theorem}

\begin{proof}
Define $\Omega = \{(\emptyset, F)\}$. Then the nondeterministic Rabin automaton $A'= (Q, \Sigma, q_I, \delta, \Omega)$ has the desired property.
\end{proof}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\bibliographystyle{abbrv}%splncs
%\bibliography{bibliography}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\end{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%